Query Optimization for Dynamic Imputation

### 2017-05-15 paper/algorithm.tex
Pass through entire paper

\subsection{Imputation Operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query.

@@

Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$. In essence, $\alpha$ is a tunable parameter 
that determines whether the optimizer focuses on quality or on performance, as plans that
lose a lot of information through dropping are also the fastest.

If $\alpha=0.0$, then the optimal query should lose as little information as possible (at
the expense of performance). If $\alpha = 1.0$, then the optimal query should have the
lowest runtime (at the expense of quality).

@@

\noindent{\textbf{Constructing $P$}}: The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations of distribution statistics such as mean and variance.
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
Finally, the penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

\noindent{\textbf{Constructing $T$}}:
As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
If we assume that the \textit{Impute} routine is CPU-bound, then we adjust $T$ by some constant factor $f$, thereby scaling the result to be comparable with IO-bound routines.

@@

We note here that for every imputation operator inserted into the query plan, a new
statistical model is instantiated and trained. Though it is tempting to try to pre-train
imputation models while the analyst begins to make queries, the exact rows of data, set
of complete attributes, and set of attributes to impute are unknown until runtime, making
this strategy infeasible. Thus, if applicable, $T$ should take into account both
imputation model training and application.

We discuss several values of $T$ for different imputation methods in \Cref{sec:imputation}.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, by supplying penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These cost functions are then used by the optimizer during planning.
The flexibility of this approach is in the spirit of query optimization, which aims to reconcile performance with a declarative interface to data manipulation.
To motivate this API, we describe how a developer would specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target attributes. 
In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

\begin{equation*}
\begin{aligned}
    T(q) &\approx 0 \\
    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
\end{aligned}
\end{equation*}

@@

\begin{equation*}
\begin{aligned}
    T(q) &= \textsc{Card}(q) \\
    P(q) &= \sum_{A \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(A)}{\textsc{Card}(q)}
\end{aligned}
\end{equation*}
}

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer is the set of plans with the best imputation quality-execution time trade-offs.

\subsubsection{Planning algorithm}
The planner (\Cref{algo:top-level-planner}) operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

@@

To motivate the restriction of imputation types, we consider the implications of allowing arbitrary imputations.
If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator where $D$ is the set of dirty attributes in all tables.
This would increase the overall complexity of the algorithm drastically.

>>>>

\subsection{Imputation Operators}
\label{sec:operators}
We introduce two new relational operators to perform imputation: \textit{Impute} ($\mu$) and
\textit{Drop} ($\delta$). Each operator takes arguments $(C, R)$ where $C$ is a set of
attributes and $R$ is a relation. \textit{Impute} uses a machine learning algorithm to
replace all \nullv{} values with non-\nullv{} values for attributes $C$ in the relation $R$ (discussed in detail in~\Cref{sec:imputation}).
% In our experiments, we implement \textit{Impute} using a general-purpose chained-equation decision trees algorithm, but our system is agnostic to this choice.
\textit{Drop} simply removes from $R$ all tuples which have a \nullv{} value for some attribute in $C$.
Both operators guarantee that the resulting relation will contain no \nullv{} values for
attributes in $C$.

@@

This set $Q$ will be the Pareto frontier~\cite{pareto1964cours} of $S$, and it contains the best option for all feasible trade-offs of $\textsc{Time}$ and $\textsc{Penalty}$ for the current query. This
set of plans is significantly smaller than the set of all plans, which our algorithm prunes away.

@@

Given $\alpha$ and $Q$, \ProjectName{} returns an $\alpha$-bound optimal plan in $Q$. In essence, $\alpha$ is a tunable parameter 
that determines whether the optimizer focuses on quality or on performance, as plans that
lose a lot of information through dropping are also the fastest. 
If $\alpha=0.0$, then the optimal query should lose as little information as possible (at
the expense of performance). If $\alpha = 1.0$, then the optimal query should have the
lowest runtime (at the expense of quality).

@@

\textbf{Constructing $P$}:
The form of $P$ is chosen to reflect the properties of the imputation method being used.
A well-chosen $P$ should provide a reasonable estimate of the quality of imputation from the set of plan histograms.
From these, $P$ can obtain the number of clean and dirty attributes in its input, their cardinalities, and approximations of distribution statistics such as mean and variance.
Intuitively, imputation quality \textit{increases} when more complete attributes and complete tuples are available; correspondingly, $P(q')$ \textit{decreases} as the number of values increases.
The penalty should also reflect the intuition that the number of training observations has diminishing returns by decreasing more slowly as the number of values increases.
Finally, the penalty may increase as amount of noise or spread in the data increases, as this makes it more difficult to obtain precise imputations.

@@

\textbf{Constructing $T$}:
As with $\textsc{Penalty}$, the time computation for \textit{Impute} depends on the properties of the underlying algorithm via a function $T$.
If we assume that the \textit{Impute} routine is CPU-bound, then we adjust $T$ by some constant factor $f$, thereby scaling the result to be comparable with IO-bound routines.

@@

We note here that for every imputation operator inserted into the query plan, a new
statistical model is instantiated and trained. Though it is tempting to try to pre-train
imputation models while the analyst begins to make queries, the exact rows of data, set
of complete attributes, and set of attributes to impute are unknown until runtime, making
this strategy infeasible. Thus, if applicable, $T$ should take into account both
training and application of the imputation model.

@@

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, by supplying penalty ($P$) and time ($T$) functions (\Cref{sec:cost-model}) pertinent to the algorithm.
These cost functions are then used by the optimizer during planning.
The flexibility of this approach is in the spirit of query optimization, which aims to reconcile performance with a declarative interface to data manipulation.
To motivate this API, we describe how a developer might specify these functions for a general-purpose imputation algorithm, as well as for two simpler imputation methods.

@@

The chained-equation algorithm proceeds by
iteratively fitting decision trees for the target attributes. 
In each iteration, the missing values of a single attribute are replaced with newly imputed values.  Values in a single column can be updated over multiple
iterations of the algorithm. With more iterations, the quality of the
imputation improves as values progressively reflect more accurate relationships
between attributes. The algorithm terminates after a fixed number of
iterations, or earlier if convergence is achieved, indicated by unchanged values in a 
column in subsequent iterations. For our experiments, we use Weka's \texttt{REPTree}
decision tree implementation~\cite{witten2016data}, a variant of the canonical \texttt{C4.5}
algorithm~\cite{quinlan1993}. Roughly, the time complexity of building one
decision tree is $\mathcal{O}(nm \text{ } log(n))$, where $n$ is the number of tuples
and $m$ is the number of dependent attributes~\cite{witten2016data}. Since we build
$k * |\textsc{Dirty}(q)|$ trees, we have

@@

\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &\approx 0
\end{aligned}
\end{equation*}

@@

\begin{equation*}
\begin{aligned}
    P(q) &= \sum_{c \in \textsc{Dirty}(q)} \frac{1+\widehat{Var}(c)}{\textsc{Card}(q)} \\
    T(q) &= \textsc{Card}(q)
\end{aligned}
\end{equation*}
}

@@

The plan cache uses the partial order of plans defined by  $\langle \textsc{Dirty}, \textsc{Penalty}, \textsc{Time} \rangle$ to
collect sound and complete Pareto frontiers. Plans with different $\textsc{Dirty}$ sets cannot be compared. The final
Pareto frontier produced by the optimizer is the set of plans with the best imputation
quality and runtime trade-off.

\subsubsection{Planning algorithm}
The planner (\Cref{algo:top-level-planner}) operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections
are pushed to the leaves, and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

@@

To motivate the restriction of imputation types, we consider the implications of allowing arbitrary imputations.
If we allow any arbitrary subset of attributes to be imputed, then we would need to consider $O(2^{|D|+1})$ different imputation operators before each relational operator where $D$ is the set of dirty attributes in all tables.
This would increase the overall complexity of the algorithm dramatically.

### 2017-05-15 paper/examples.tex
Pass through entire paper

\subsection{Planning with \ProjectName{}}
The search space contains plans of varying performance and quality of results, as a product of the multiple possible locations for imputation.
The user can influence the final plan selected by \ProjectName{}
through a trade-off parameter $\alpha \in [0, 1]$, where low $\alpha$ 
prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

@@

\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operation $\mu$,
which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, though other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

@@

An alternative strategy involves starting with $\alpha=1$ and iterating to lower values, as long as execution
of the query completes within some user-defined time bound. \ProjectName{}'s planning
algorithm can expose the final set of plans that tradeoff execution time and imputation quality, which allows
the user to adjust their choice of $\alpha$ to uniquely select amongst these plans.

Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
Knowing that there may be a need to explore this data set further, she can easily incorporate her imputation model into ImputeDB for future use.

>>>>

\subsection{Planning with \ProjectName{}}
The search space contains plans with varying performance and imputation quality
characteristics, as a product of the multiple possible locations for imputation.
The user can influence the final plan selected by \ProjectName{}
through a trade-off parameter $\alpha \in [0, 1]$, where low $\alpha$ 
prioritizes the quality of the query results and high $\alpha$ prioritizes performance.

@@

\Cref{fig:quality-plan} shows a quality-optimized plan that uses the impute operator $\mu$,
which employs a reference imputation strategy to fill in missing values rather than dropping tuples.
It waits to impute \verb|demo.income| until after the final join has taken place, though other imputations take place earlier on in the plan, some before filtering and join operations.
Imputations are placed to maximize ImputeDB's estimate of the quality of the overall results.
Meanwhile, \Cref{fig:fast-plan}'s performance-optimized plan uses the drop operation $\delta$ instead of filling in missing values.
However, it is a significant improvement over dropping all tuples with missing data in any field, as it only drops tuples with missing values in attributes which are referenced by the rest of the plan. In both cases, only performing
imputation on the necessary data yields a query that is much faster
than performing imputation on the whole dataset and then executing
the query.

@@

\review{
As an alternative strategy, the analyst can start by executing her query with $\alpha=1$ and
incrementally reducing $\alpha$, as long as execution of the query completes within some
acceptable time bound.
\ProjectName{}'s planning algorithm can expose the final set of plans with optimal
time-quality trade-offs, which allows the user to adjust her choice of $\alpha$ to select from these plans.
}

Tens of queries later, our epidemiologist has a holistic view of the data and she has the information that she needs to construct a tailored imputation model for her queries of interest.
Knowing that there may be a need to explore this data set further, she can also incorporate her imputation model into ImputeDB for future use.

### 2017-05-15 paper/experiments.tex
small changes
Pass through entire paper

\section{Experiments}\label{sec:experiments}
For our experiments we plan and execute queries for three separate survey-based data sets, showing that our system is well suited for early dataset exploration.
We show that \ProjectName{} performs an order of magnitude better than traditional
base-table imputation and produces results of comparable quality.

@@

\subsubsection{CDC NHANES}
For our first set of experiments, we use survey data collected by the 
U.S. Centers for Disease Control and Prevention (CDC). We
experiment on a set of tables collected as part of the 2013--2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
The data consists on survey responses, physical examinations, and laboratory
results, amongst others.

@@

The original tables have a large number of attributes, in some cases providing more granular tests results or alternative metrics.
We focus on a subset of the attributes for each table to simplify the presentation and exploration of queries.
\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
For readability, we have replaced the NHANES attribute names with self-explanatory attribute names.

@@

\subsubsection{freeCodeCamp 2016 New Coder Survey}
For our second set of experiments, we use data collected by freeCodeCamp, an open-source
community for learning to code, as a part of a survey of new software
developers~\cite{fcc-data}.  The \textit{2016 New Coder Survey} consists of responses by
over 15,000 people to 48 different demographic and programming-related questions.  The
survey targeted users who were related to coding organizations.

@@

\subsubsection{American Community Survey}
For our final experiment, we run a simple aggregate query over data from the American
Community Survey (ACS), a comprehensive survey conducted by the U.S.
Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
data kindly provided by \citeauthor*{akande2015empirical}. Given that the data had
been cleaned, we artificially dirtied it by replacing 40\% of the values uniformly at random with \nullv{} values.
The final dataset consists of 671,153 rows and 37 integer columns.

@@

To calculate SMAPE for each $\alpha$,
we compute tuple-wise absolute percentage deviations within each iteration of each query, and
average this value over all iterations.
Relative errors of query results, when optimizing for quality
($\alpha=0$), are low --- between \lowsmapealphazero{} and
\highsmapealphazero{}\% --- indicating that on-the-fly imputation produces
results similar to the baseline. When optimizing for performance ($\alpha=1$),
relative errors can be modestly higher --- up to \highsmapealphaone{}\% in the
worst case, as this approach most closely corresponds to dropping all
\nullv{}s. Thus, it is important to recognize the tradeoffs between quality and
performance for a specific use case.

\review{
We also experiment with mean-value imputation and hot
deck imputation (\Cref{sec:imputation}). One approach to
mean-value imputation, to facilitate non-blocking operation within the iterator
model, is to estimate the column mean from base table histograms. In this case,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.
}
}

We also calculate the number of tuples used to produce each aggregate output.
The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped.
Even in cases where the SMAPE reduction between $\alpha = 1$ and $\alpha = 0$ is small (\Cref{q2}, \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data. In particular, if values are missing uniformly at random,
the aggregate should not be affected.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
\review{
By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.

\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm.
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

@@

%\subsubsection{Planning Times}
%\begin{figure}
%\centering
%\includegraphics[scale=0.4]{figures/planning_times_plot.png}
%\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets. Users can increase the level of
%approximation for more aggressive planning reductions.
%}
%\label{fig:planning-times}
%\end{figure}

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

>>>>

\section{Experiments}\label{sec:experiments}
For our experiments we plan and execute queries for three separate survey-based data sets.
We show that \ProjectName{} performs an order of magnitude better than traditional
base-table imputation and produces results of comparable quality, showing that our system is well suited for early dataset exploration.

@@

\subsubsection{CDC NHANES}
For our first set of experiments, we use survey data collected by the 
U.S. Centers for Disease Control and Prevention (CDC). We
experiment on a set of tables collected as part of the 2013--2014 National
Health and Nutrition Examination Survey (NHANES), a series of studies
conducted by the CDC on a national sample of several thousand individuals~\cite{cdc-data}.
The data consists of survey responses, physical examinations, and laboratory
results, among others.

@@

The original tables have a large number of attributes, in some cases providing more granular test results or alternative metrics.
We focus on a subset of the attributes for each table to simplify the presentation and exploration of queries.
\Cref{table:nhanes-description} shows the attributes selected, along with the percentage of \nullv{} values for each attribute.
For readability, we have replaced the NHANES attribute names with self-explanatory attribute names.

@@

\subsubsection{freeCodeCamp 2016 New Coder Survey}
For our second set of experiments, we use data collected by freeCodeCamp, an open-source
community for learning to code, as part of a survey of new software
developers~\cite{fcc-data}.  The \textit{2016 New Coder Survey} consists of responses by
over 15,000 people to 48 different demographic and programming-related questions.  The
survey targeted users who were related to coding organizations.

@@

\subsubsection{American Community Survey}
For our final experiment, we run a simple aggregate query over data from the American
Community Survey (ACS), a comprehensive survey conducted by the U.S.
Census Bureau. We use a cleaned version of the 2012 Public Use Microdata Sample (PUMS)
data, which we then artificially dirty by replacing 40\% of the values uniformly at random with \nullv{} values.
The final dataset consists of 671,153 rows and 37 integer columns.

@@

To calculate SMAPE for each $\alpha$,
we compute tuple-wise absolute percentage deviations within each iteration of each query, and
average this value over all iterations.
Relative errors of query results, when optimizing for quality
($\alpha=0$), are low --- between \lowsmapealphazero{} and
\highsmapealphazero{}\% --- indicating that on-the-fly imputation produces
results similar to the baseline. When optimizing for performance ($\alpha=1$),
relative errors can be modestly higher --- up to \highsmapealphaone{}\% in the
worst case, as this approach most closely corresponds to dropping all
\nullv{}s. Thus, it is important to recognize the trade-offs between quality and
performance for a specific use case.

\review{
We also experiment with mean-value imputation and hot
deck imputation (\Cref{sec:imputation}). In our implementation of mean-value imputation,
to facilitate non-blocking operation within the iterator
model, we estimate the column mean from base table histograms. Therefore,
imputations are identical no matter where the operator is placed in the query
plan. Moreover, runtimes are essentially identical regardless of the value of
$\alpha$ because the time cost of the imputation is relatively low no matter
how many tuples need to be imputed. In hot deck imputation, runtimes are also
similar across $\alpha$, because the cost of buffering
tuples and sampling at random is negligible compared to the rest of query
execution. \todo{be quantitative: In terms of accuracy, relative errors using hot deck are close to
zero; since the baseline, in this case, is hot deck imputation on the base
table, only skew in the distributions arising from filters and joins would lead
to larger errors. In some cases, skew does arise, and errors are on the same
magnitude as in the decision trees case. Indeed, \ProjectName{} yields larger
benefit when using relatively sophisticated, higher quality imputation
algorithms as opposed to simpler strategies that would even be quick to
implement on the base table.
}
}

We also calculate the number of tuples used to produce each aggregate output.
The count fraction columns in \Cref{table:smape} show the number of tuples in the aggregate for $\alpha = 0$ and $\alpha = 1$ as a fraction of the number of tuples used when running the query on the imputed base table.
This shows that when optimizing for performance, not quality, many tuples are dropped due to
insertion of $\lambda$ operators.
Even in cases where the SMAPE reduction from $\alpha = 1$ and $\alpha = 0$ is small
(\Cref{q2} and \Cref{q6}) the tuple count is significantly different.
In these cases, the aggregate value is not significantly impacted by the missing data. In
particular, if values are missing completely at random,
the aggregate should not be affected.
However, if the missing data is biased then the aggregate will have a significant error.
This highlights a challenge for a user handling data imputation traditionally: it is unclear if the missing data will have a large or small negative impact on their analysis until they have paid the cost of running it.
\review{
By using \ProjectName{} this cost can be lowered significantly. We can also trivially extend \ProjectName{} to warn users when the query chosen for execution has a high
$\textsc{Penalty}$ estimate, along with the number of tuples that have been dropped (after execution), so that situations with high potential for skewed results can be identified by the user.

\Cref{fig:pareto-frontiers} shows the final Pareto frontiers produced by the planner for \Cref{q1} and \Cref{q6}. These are the resulting set of plans after pruning dominated plans throughout our algorithm
and encapsulate the possible trade-offs that the user can make in choosing a final plan for execution. The frontier is available at the end of the planning stage and
can be exposed to the user to help in guiding their workflow.
}

@@

\review{
However, \ProjectName{}'s planning algorithm is still exponential in the number of joins in the plan, and the use of Pareto sets exacerbates this exponential growth.
We explored the limits of our planning algorithm by constructing a series of queries with increasing number of joins over the CDC tables.
Planning times for queries involving 1--5 joins---a practical value for  real world exploratory queries---are less than 1 second.
We extended \ProjectName{} to support an approximation of Pareto sets to improve planning times for queries with 6 or more joins. Approximate Pareto sets
displayed a reduction in planning time linear with respect to the reduction in size of the final frontier. For queries with 6--8 joins, approximate sets achieved a reduction
in planning time between 14\% and 23\%, on average. Users can increase the level of approximation to reduce planning times further.
}

@@

%\subsubsection{Planning Times}
%\begin{figure}
%\centering
%\includegraphics[scale=0.4]{figures/planning_times_plot.png}
%\caption{Planning times are sub-second for up to 5 joins. For more joins, we implement an approximation to Pareto sets. Users can increase the level of
%approximation for more aggressive planning reductions.
%}
%\label{fig:planning-times}
%\end{figure}
%\todobox{Jose: I'm not a fan of this line, I would remove}{Furthermore, at that point, it may make more sense to specify a coherent model over the entire base table instead.}

### 2017-05-15 paper/implementation.tex
Pass through entire paper

\begin{itemize}
\item \textbf{Extended histograms:}
  Ranking query plans with imputation needs estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
  We extended standard histograms to include the count of missing values for each attribute.
  On the base tables, this provides an exact count of the missing values.
  In sub-plans we use simple heuristics---as discussed in Section~\ref{sec:cardinal}---to estimate the number of missing values after applying both standard and imputation operators.
  These cardinality estimates are critical to the optimizer's ability to compare the performance and imputation quality of different plans.

@@

\item \textbf{Integrated optimizer:}
  We believe that it would be relatively simple to extend an existing query optimizer to insert imputations immediately after scanning the base tables.
  However, integrating the imputation placement rules into the optimizer allows us to explore a large space of query plans which contain imputations.
  In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
\end{itemize}
}

>>>>

\begin{itemize}
\item \textbf{Extended histograms:}
  Ranking query plans that include imputation requires estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
  We extend standard histograms to include the count of missing values for each attribute.
  On the base tables, this provides an exact count of the missing values.
  In sub-plans we use simple heuristics---as discussed in Section~\ref{sec:cardinal}---to estimate the number of missing values after applying both standard and imputation operators.
  These cardinality estimates are critical to the optimizer's ability to compare the performance and imputation quality of different plans.

@@

\item \textbf{Integrated optimizer:}
  We believe that it would be relatively simple to extend an existing query optimizer to insert imputations immediately after scanning the base tables.
  However, integrating the imputation placement rules into the optimizer allows us to explore a large space of query plans which contain imputations.
  In particular, this tight integration allows us to jointly choose the most effective join order and imputation placement.
  Our goal of integrating imputation placement into the optimizer requires the other modifications that we have discussed.
\end{itemize}
}

### 2017-05-15 paper/introduction.tex
Pass through entire paper

Many databases have large numbers of missing or \nullv{} values;  these can arise for a variety of reasons, including missing source data (e.g., unknown facts),
missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

@@

A simpler approach might drop all rows with missing values, which can not only introduce
bias into the results, but also result in discarding much of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
carefully plans the placement of imputation/row-drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the
Pareto-optimal frontier of plans that trade imputation cost for result quality,
and allow the analyst to specify their desired plan from this frontier.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazero{} times as fast as it takes to impute all missing values and then run queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaone{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).

>>>>

Many databases have large numbers of missing or \nullv{} values;  these can arise for a
variety of reasons, including missing source data,
missing columns during data integration, de-normalized databases, or as a result of outlier detection and cleaning where \nullv{} values are substituted for
bad values~\cite{kim2003}.  Such \nullv{} values can lead to incorrect or ill-defined query results~\cite{rubin1976}, and as such removing these values from data
before processing is often desirable.

@@

A simpler approach might drop all rows with any missing values, which can not only introduce
bias into the results, but also result in discarding much of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set. Specifically, \ProjectName{}
carefully plans the placement of imputation or row-drop operations inside the query plan, resulting in a 
significant speedup in query execution while reducing the number of dropped rows. Unlike previous work, the focus of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the
Pareto-optimal frontier of plans that trade imputation cost for result quality,
and allow the analyst to specify their desired plan from this frontier.

@@

% Implications
Our approach enables an exploratory analysis workflow in which the analyst can issue
standard SQL queries over a data set, even if that data has missing values.
These queries execute between \lowxalphazero{} and \highxalphazero{} times faster than
imputing all missing values and then running queries (the traditional approach). Furthermore, the results obtained with this approach
are similar in quality to those obtained with the traditional imputation approach
--- between \lowsmapealphazero{} and \highsmapealphaone{}
percent, in one error measure --- in our empirical results over real-world data sets (see~\Cref{sec:experiments}
for details).
